---
title: "Stochastic Block Models for Multiplex  networks"
subtitle: "Principle and illustration"
author: "team großBM"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
bibliography: references.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Multiplex Stochastic Block Models}
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Preliminaries

This vignette illustrates the use of the `estimateMultiplexSBM` function and the methods accompanying the R6 classes `multiplexSBMfit`.

### Requirements

The packages required for the analysis are **sbm**. 

```{r setup, message=FALSE, warning=FALSE}
library(sbm)
```



### Mathematical Background

We give here a few details on Stochastic Block Models for multiplex networks. 
Multiplex networks refer to a collection of networks involving the same nodes, each network corresponding to a give type of interaction. 
Such a network is studied in  @multiplexSBM where the nodes are researchers and two networks are provided: 
    
    * a network representing the relations of advices between researchers,
    * a network representing the exchanges between the laboratories the researchers belong to. 

In @multiplexKefi, the nodes are species and each network corresponds to a type of ecological relation, namely mutualism, competition and **** 


See @multiplexSBM for details.

### Inference 

The model selection and the estimation are performed with the function `estimatemultipartiteBM`.

```{r load result, echo = FALSE, eval = TRUE}
load('resMultipartiteEcological.rda')
``` 

```{r MBM, echo = TRUE, eval = FALSE}
estimOptions = list(initBM = FALSE)
listSBM <- list(PlantFlovis, PlantAnt, PlantBird)
myMSBM <- estimateMultipartiteSBM(listSBM,estimOptions)
``` 


\code{myMSBM} contains the estimated parameters of the models we run through during the search of the better numbers of blocks. 

RES_MBM contains the dataset and the results.  
```{r MBM what}
myMSBM
``` 


The best model has the following numbers of blocks
```{r MBM v_K }
myMSBM$nbBlocks
```


To see the parameters estimated for the better model we use the following command `myMSBM$connectParam` or `myMSBM$blockProp`
```{r MBM param }
myMSBM$blockProp
myMSBM$connectParam
```


The clustering  supplied by the better model  are in `myMSBM$memberships***`.

```{r MBM Z }
table(myMSBM$memberships$Plants)
table(myMSBM$memberships$Ants)      
``` 

```{r storedmodels}
myMSBM$storedModels
``` 

## Plots 
We can either plot the reorganized matrix
```{r plot, eval = TRUE }
plot(myMSBM) 
``` 

or the mesoscopic view
```{r plot meso, eval = FALSE}
plotOptions=list(vertex.size = c(6,3,2,2))
plotOptions$vertex.shape = rep('circle',4)
plotOptions$vertex.color = c('darkolivegreen3','darkgoldenrod2','salmon2','cadetblue2')
plotOptions$edge.curved = 0.1
plot(myMSBM,type = 'meso',plotOptions=plotOptions)
```

Will soon be ready

## References

