---
title: "Simple and Bipartite Stochastic Block Models"
subtitle: "An illustration on antogonistic tree/fungus network"
author: "team großBM"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
bibliography: references.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{SBM_tree_fungus_network}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Preliminaries

This vignette illustrates the use of the `estimateSBM` function and the methods accompanying the R6 classes `SimpleSBMfit` and `BipartiteSBMfit`. 

### Requirements

The packages required for the analysis are **sbm** plus some others for data manipulation and representation:

```{r setup, message=FALSE, warning=FALSE}
library(sbm)
library(ggplot2)
theme_set(theme_bw())
library(igraph)
library(alluvial)
```

### Data set: antagonistic tree/fungus interaction network

We consider the fungus-tree interazction network studied by @tree_fungus_network, available with the package **sbm**:

```{r import dataset}
data("fungus_tree_network")
str(fungus_tree_network, max.level = 1)
```

This data set provides information about $154$ fungi sampled on $51$ tree species. It is a list with the following entries:

  - `fungi_list`: list of the fungus species names
  - `tree_list` : list of the tree species names
  - `fungus_tree` : binary fungus-tree interactions
  - `tree_tree` : weighted tree-tree interactions (number of common fungal species two tree species host) 
  - `covar_tree` : covariates associated to pairs of trees (namely genetic, taxonomic and geographic distances)

### Mathematical Background

See @blockmodels for details.

## Analysis of the tree/fungus data

### Tree-tree binary interaction networks

```{r tree_tree_binary network}
tree_tree_binary <- 1 * (fungus_tree_network$tree_tree != 0)
```

```{r tree_tree_binary network plot data}
corrplot::corrplot(tree_tree_binary, is.corr = FALSE, method = "color", cl.pos = "n")
```

```{r simpleSBM}
mySimpleSBM <- tree_tree_binary %>% 
  estimateSimpleSBM("bernoulli", estimOptions = list(verbosity = 0))
```

```{r}
mySimpleSBM$nbBlocks
mySimpleSBM$plot(type = 'expected')
```
```{r simpleSBM ICL}
mySimpleSBM$storedModels %>% 
  ggplot() + aes(x = nbBlocks, y = ICL) + geom_line() + geom_point(alpha = 0.5)
```
```{r changeModel}
mySimpleSBM$setModel(4)
mySimpleSBM$nbBlocks
mySimpleSBM$plot(type = 'expected')
```


### Analysis of weigthed interaction network


## References

